---
title: "statistical_project"
output: pdf_document
date: "2024-05-27"
Authors: Daniel Gutierrez, Rafael Hernandez, Camilo Sanabria
---

# Project Introduction

In this project we aim to study the influence of macroeconomics indicators on the house price index in Colombia. This indicator measure the evolution through time of the market prices of residential properties as a percentage change. The idea of the analysis is to present and study the possible effects of the macroeconomics indicators on the house prices through the construction of regression models, evaluating their results and interpreting the models in the context of the data.

# Dataset Introduction and Preprocessing

The dataset was constructed gattering information from multiple sites: the Colombian department of statistics (DANE) , the Colombian central bank (Banco de la Repúlbica), Google trends and the Federal Reserve Economic Data (FRED). In general, all the visited sites present the data as a .csv file with tables according to filters related with the time span of interest. With this we construct a consolidated database in .csv format with the following variables:

| Variable                   | Description                                                                                                    |
|------------------------|------------------------------------------------|
| House Price Index       | Measure of the average change in the residential properties price as an index                                  |
| Industrial Inputs Index          | Measure of the average changes in the industry input costs as an index                                         |
| Metals Price Index                     | Measure of the average price change in metal costs excluding gold as an index                                  |
| Energy Price Index                     | Measure of the average price change of energy including Crude oil, Natural Gas, Coal Price and Propane Indices |
| Shipping Price Index                   | Measure the average price change of shipping costs                                                             |
| Forex Index                         | Indicator of the foreign exchange rate of the Colombian Peso (COP) with respect of the USD                     |
| Unemployment Rate              | Indicator of the percentage of unemployment population in Colombia                                             |
| Industrial Production Index           | Measure of the level of production on industrial sectors as an index                                           |
| Interest rate              | Reference interest rate emited by the colombian central bank with respect of the others financial institutions |
| Construction Licences Area | Measure the total of squared meters given for construction licences in Colombia                                |
| Finished Constructions     | Measure of the total constructions that have been finished in Colombia                                         |
| Google Trends Housing              | Google search trends on housing for Colombia                                                                   |

These variables present real values,and were selected as they have an initial coherent relation with the housing sector. We present the import process of the data
```{r}
#`echo = FALSE`: Hides the R code in the final document.
#`results = 'hide'`: Hides the output of the R code.
```

```{r import-data, echo=FALSE,message=FALSE, warning=FALSE}
# Necessary libraries to run all the code chunks and plots
library(readxl)
library(dplyr) 
library(car) 
library(lmtest) 
library(zoo)
library(caret)
```

# Data Uploading

```{r}
#Read the data using the read_excel function
#Change here for current data path
data <- read_excel("C:/Users/CAMILO/Documents/GitHub/unipd_sl_24/data/data.xlsx",sheet = "dataframe_col")
#View(data)
#Present the first 5 rows of the data
#head(data,5)
```

# Data pre-processing and cleaning

After have an initial version of the data we proceed with the pre-processing and cleaning steps. As a first approximation we identify the number of NA values in each one of the columns.

```{r}

#We first change the variables names to have more consistency 

colnames(data) <- c('MY','date','Home_Price_Index','Industrial_Inputs_Index','Metals_Price_Index','Energy_Price_Index','Shipping_Price_Index','Forex_Index','Unemployment_Rate','Industrial_Production_Index','Interest_Rate','Construction_Licences_Area','Finished_Constructions','Google_Trends_Housing') 
# Counting NA values in each column
na_counts <- apply(data, 2, function(x) sum(is.na(x)))
# Print the counts of NA values per column
print(na_counts)

# Construction and Industrial prod have nans, 
# they are at the extremes 
# Convert date column to Date type if it's not already
data$date <- as.Date(data$date)

# Copy the dataframe to avoid modifying the original one
data_filled <- data
```
We realize that the variable finished_constructions requires an additional pre-processing step because it is recorded by quarters while the other are registered by months. We propose the following reconstruction steps:

```{r}
# Interpolate finished_constructions
# Find the indices where the date is at the end of a trimester
trimester_end_indices <- which(format(data$date, "%m") %in% c("03", "06", "09", "12"))

# Loop through each trimester end and distribute the value to the previous three months
for (i in trimester_end_indices) {
  if (i - 2 > 0) {
    # Distribute the value to the current month and the previous two months
    value_to_distribute <- data$Finished_Constructions[i] / 3
    data_filled$Finished_Constructions[i] <- value_to_distribute
    data_filled$Finished_Constructions[i - 1] <- value_to_distribute
    data_filled$Finished_Constructions[i - 2] <- value_to_distribute
  }
}

data_filled$Finished_Constructions <- na.locf(data_filled$Finished_Constructions, na.rm = FALSE)

#View(data_filled)
```


# Summary stats and final dataset

We proceed to remove the left rows with NA values eliminating from the dataset the registers before January 2005, also we drop the column MY also referred to the date in form of MONTH()YEAR(). Finally, we present a summary of the dataset after these transformations. In particular we see that the Housing Price Index range present values on [87.35,203.50] with an IQR=70.38.

```{r}
# Removing rows with any NA values
clean_data <- na.omit(data_filled)

# Drop columns ------------

data_reduced <- subset(clean_data, select = -c(MY))
#head(data_reduced,10)
```


```{r}
# View description --------------
# Get the summary of the dataframe
summary_stats <- summary(select(data_reduced,-date))

# Print the summary statistics
#print(summary_stats)


```


| Statistic    | Min    | 1st Qu. | Median | Mean   | 3rd Qu. | Max    |
|--------------|--------|---------|--------|--------|---------|--------|
| Home_Price_Index            | 87.35  | 110.50  | 157.92  | 147.71 | 180.88  | 203.50 |
| Industrial_Inputs_Index     | 78.67  | 116.02  | 136.26  | 137.78 | 157.65  | 221.14 |
| Metals_Price_Index          | 74.89  | 121.21  | 143.74  | 145.88 | 172.53  | 234.53 |
| Energy_Price_Index          | 55.89  | 128.19  | 162.61  | 173.36 | 223.35  | 376.41 |
| Shipping_Price_Index        | -213.3 | 242.2   | 262.1   | 280.7  | 294.1   | 478.3  |
| Forex_Index                 | 0.1712 | 1.1934  | 2.3762  | 2.6762 | 3.2021  | 4.9223 |
| Unemployment_Rate           | 0.07563| 0.09728 | 0.10970 | 0.11240| 0.12099 | 0.21972|
| Industrial_Production_Index | 59.82  | 77.94   | 94.44   | 92.87  | 104.85  | 132.46 |
| Interest_Rate               | 0.01750| 0.03860 | 0.04525 | 0.05551| 0.06994 | 0.13250|
| Construction_Licences_Area  | 233210 | 969559  | 1156424 | 1205587| 1373585 | 2883878|
| Finished_Constructions      | 471310 | 775374  | 848663  | 871488 | 962168  | 1335811|
| Google_Trends_Housing       | 16.00  | 23.50   | 34.75   | 36.04  | 47.62   | 64.00  |

# Exploratory Analysis


# Outliers based on the IQR

As an initial step to better understand the data and its behavior we find the outliers of each one of the variables using the IQR to select those points outside the bounds. As a result we found that the variables that present the most quantity of outliers are the Shipping Price Index, the Interest Rate and the Unemployment Rate.

```{r}
# Function to detect outliers based on IQR
detect_outliers_single_var <- function(column) {
  column <- na.omit(column) 
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  outliers <- column[column < lower_bound | column > upper_bound]
  return(outliers)
}

# List to store outliers for each variable
outliers_list <- list()

# Loop through each column in the dataset
for (var in colnames(select(data_reduced,-date))) {
  if (is.numeric(data[[var]])) {
    outliers_list[[var]] <- detect_outliers_single_var(data[[var]])
  }
}

# Print outliers for each variable
print(outliers_list)


```
| Variable                      | Values                                                                                                     |
|-------------------------------|------------------------------------------------------------------------------------------------------------|
| Home_Price_Index              | 0                                                                                                |
| Industrial_Inputs_Index       | 0                                                                                                |
| Metals_Price_Index            | 0                                                                                                |
| Energy_Price_Index            | 376.4121                                                                                                  |
| Shipping_Price_Index          | 374.969, 375.187, 395.268, 446.983, 439.181, 454.407, 468.721, 478.263, 478.216, 471.877, 469.290, 442.591, 456.584, 412.082, 409.599, 443.799, 420.694, 446.214, 462.462  |
| Forex_Index                   | 0                                                                                                |
| Unemployment_Rate             | 0.1725640, 0.2048530, 0.2197200, 0.2035910, 0.2091470, 0.1744270, 0.1629280, 0.1756237                    |
| Industrial_Production_Index   | 0                                                                                                |
| Interest_Rate                 | 0.1141935, 0.1204839, 0.1275000, 0.1275806, 0.1300000, 0.1324194, 0.1325000, 0.1325000                    |
| Construction_Licences_Area    | 2493085, 1988396, 2838778, 2337743, 2331210, 2859035, 2372414, 2052607                                    |
| Finished_Constructions        | 4007432, 3762696, 1413931                                                                                 |
| Google_Trends_Housing         | 0        

In addition present the box plot for the variables with the highest quantity of outliers

```{r,echo = FALSE,fig.width=3, fig.height=5,fig.align='center'}
# Create the boxplot without x-axis labels
# Adjusting the margins: bottom, left, top, right
{
par(mfrow = c(3, 1), mar = c(5, 4, 2, 2) + 0.1, oma = c(0, 0, 0, 0))
boxplot(data_reduced$Shipping_Price_Index, horizontal=TRUE  ,frame=F,main = "Shipping Price Index")

boxplot(data_reduced$Unemployment_Rate, horizontal=TRUE  ,frame=F,main = "Unemployment Rate")

boxplot(data_reduced$Interest_Rate, horizontal=TRUE  ,frame=F,main = "Interest Rate")
}
```

# Ploting the Home Price Index

Continuing with the exploratory analysis we plot the variable of Home Price Index to see its behavior through time. It is possible to see a clear increase tendency starting from the year 2005 until the year 2021, then a little decay was recorded.

```{r,echo = FALSE,fig.width=6, fig.height=4,fig.align='center'}
# reset margins
# Reset margins to default values
par(mar = c(5, 4, 4, 2) + 0.1)
par(mfrow = c(1, 1))

# dependent variable
data_reduced$category <- as.factor(clean_data$date)
# Set up the plot without drawing the points
{plot(1:length(data_reduced$category), clean_data$Home_Price_Index, type = "n", xaxt = "n",xlab="date", ylab = "Home Price Index", main = "Evolution of Home Prices over time",cex.axis=0.8)

# Add the categorical labels to the x-axis
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)

# Add the lines and points to the plot
lines(1:length(data_reduced$category), data_reduced$Home_Price_Index, type = "o", col = "blue", pch = 16)}
```

# Ploting percentual variation of the Home Price Index

Another useful information can be found in the box plots and histograms for the percentual variations. We take the percentual variation of the variables with the intention of identify patterns and tendencies in a clearly way, considering that the majority of it is composed by index, and the study of it changes is fundamental for a complete understanding of their behavior throught time. We start with the Home Price Index variable.

In these plots we see that the central value of the percentual variations is positive, indicating and overall increase in the Home Price Index through time. Also the negatives outliers were expected because those negatives variations due to the decrease after the 2021.In the histogram it is possible to see a right skewed behavior.

```{r}
# Create Differences (Inflation)

# Select subset of variables to difference
variables_to_calculate <- c("Home_Price_Index", "Industrial_Inputs_Index", "Metals_Price_Index",
                            "Energy_Price_Index", "Shipping_Price_Index", "Forex_Index",
                            "Industrial_Production_Index", "Construction_Licences_Area",
                            "Finished_Constructions", "Google_Trends_Housing")

# Create a function to calculate the 12-month percentual variation
percentual_variation_12_months <- function(x) {
  return((x / lag(x, n = 12) - 1))
}

# Helper function to lag data
lag <- function(x, n) {
  c(rep(NA, n), head(x, -n))
}

# Initialize the result data frame with the original data
data_percentual_variation <- data

# Apply the function to the subset of variables
for (var in variables_to_calculate) {
  new_var_name <- paste0("pct_var_", var)
  data_percentual_variation[[new_var_name]] <- percentual_variation_12_months(data[[var]])
}

```

```{r,echo = FALSE,fig.width=9, fig.height=4,fig.align='center'}
# Histograms/Boxplots of the percentual variations
# Subset only percentual changes
columns_to_select <- c("pct_var_Home_Price_Index", "pct_var_Industrial_Inputs_Index", "pct_var_Metals_Price_Index", 
                       "pct_var_Energy_Price_Index", "pct_var_Shipping_Price_Index", "pct_var_Forex_Index", 
                       "pct_var_Industrial_Production_Index", "pct_var_Construction_Licences_Area", 
                       "pct_var_Finished_Constructions", "pct_var_Google_Trends_Housing", 
                       "Unemployment_Rate", "Interest_Rate")

data_percentual_variation <- data_percentual_variation[, columns_to_select]

data_percentual_variation$time_index<- seq(1,nrow(data_percentual_variation))


# Y
{
# Set the plot parameters
par(mar = c(5, 4, 4, 2) + 0.1)
par(mfrow = c(1, 2))

# Boxplot with color scale for time index
boxplot(data_percentual_variation$pct_var_Home_Price_Index, horizontal = TRUE, frame = FALSE, 
        main = "Boxplot of Percentual Variation Home Price Index", 
        ylab = "Percentual Variation", xlab = "", col = "lightgray", border = "black")

# Generate color scale based on time index
colors <- colorRampPalette(c("green", "red"))(length(data_percentual_variation$time_index))
colored_points <- colors[rank(data_percentual_variation$time_index)]

# Overlay points on the boxplot
points(data_percentual_variation$pct_var_Home_Price_Index, rep(1, nrow(data_percentual_variation)), 
       col = colored_points, pch = 19)

# Add a legend for the color scale
legend("topright", legend = c("Oldest", "Newest"), 
       fill = c(colors[1], colors[length(colors)]), title = "Time Index")

# Histogram of Percentual Variation Home Price Index
hist(data_percentual_variation$pct_var_Home_Price_Index, breaks = 40, 
     main = "Histogram of Percentual Variation Home Price Index", 
     xlab = "Percentual Variation")}
```

We did a similar procedure for the rest of the variables with respect to their percentual variation

```{r,echo = FALSE}

data_percentual_variation$time_index<-NULL
# Histograms-----------------
par(mar = c(2, 2, 2, 2))  # Adjust margins to make them smaller

# Adjust layout to fit all histograms
par(mfrow = c(3, 5))  # Example layout; adjust if necessary

# Histograms for each variable

for (i in 1:ncol(data_percentual_variation)) {
  hist(data_percentual_variation[[i]], main = names(data_percentual_variation)[i], xlab = names(data_percentual_variation)[i], col = "blue", border = "black")
}

```

For an easier interpretation we plot the boxplot for the percentual variation of each variable in a same graph. This plot shows that the index variables with a high variability in percentual variation are the shipping index price and the energy index price. Also, the index variable that show less variability in the percentual variation is the one related with the industrial production.

```{r,echo = FALSE,fig.width=6, fig.height=6}
# # Reset layout
# par(mfrow = c(1, 1))
# 
# #pairs(data_percentual_variation)
# 
# # Boxplot for all at once
# #install.packages("reshape2")
# library(reshape2)
# # Melt the dataframe to long format
# # Melt the dataframe to long format
# data_long <- melt(data_percentual_variation)
# 
# # Adjust graphical parameters
# 
# 
# # Create the boxplot without x-axis labels
# {par(mar = c(5, 5, 4, 2) + 0.1)  
#   boxplot(value ~ variable, data = data_long,
#         main = "Boxplot of Multiple Variables",
#         xlab = "", ylab = "Value",
#         col = "lightblue", border = "darkblue", xaxt = 'n')
# 
# # Add custom x-axis labels at a 45-degree angle
# labels <- levels(data_long$variable)
# text(x = 1:length(labels), y = par("usr")[3] - 0.5, labels = labels, srt = 45, adj = 1, xpd = TRUE, cex = 0.8)
# }
# 
# # Reset graphical parameters to default
# #par(mar = c(5, 4, 4, 2) + 0.1)
# # This one goes in the report, not so much variability in prices even a lot in X vars
# 

```

We plot these four variables to see its evolution through time. It is possible to see that the industrial production Index and the Shipping price Index have an increase tendency while the energy price have a less defined behavior over time with certain peaks in specific years like 2008 and 2022

```{r,echo = FALSE,fig.width=3, fig.height=6,fig.align='center'}
# reset margins

# dependent variable
data_reduced$category <- as.factor(clean_data$date)
# Set up the plot without drawing the points
{par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(3, 1))
  plot(1:length(data_reduced$category), clean_data$Shipping_Price_Index, type = "n", xaxt = "n",xlab="date", ylab = "Shipping Price Index", main = "Evolution of Shipping Prices Index over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Shipping_Price_Index, type = "o", col = "blue", pch = 16)

plot(1:length(data_reduced$category), clean_data$Energy_Price_Index, type = "n", xaxt = "n",xlab="date", ylab = "Energy Price Index", main = "Evolution of Energy Prices Index over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Energy_Price_Index, type = "o", col = "blue", pch = 16)

plot(1:length(data_reduced$category), clean_data$Industrial_Production_Index, type = "n", xaxt = "n",xlab="date", ylab = "Industry Production Index", main = "Evolution of Industrial Production Index over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Industrial_Production_Index, type = "o", col = "blue", pch = 16)
}
```





# Covariance and correlation

Now we present the covariance and correlation between the variables of the dataset in order to understand the strength of their linear relation. In particular we show in the following table the result for the correlation of the variables with respect to our variable y (Housing Prices Index):

| Variable                   | Correlation with Housing Prices Index |
|----------------------------|---------------------------------------|
| Industrial_inputs          | 0.1711                                |
| metals                     | 0.2371                                |
| energy                     | -0.0337                               |
| shipping                   | 0.6810                                |
| fx                         | 0.7599                                |
| unemployment               | -0.0047                               |
| industrial_prod            | 0.9418                               |
| interest_rate              | -0.2435                               |
| construction_licences_area | 0.2597                                |
| finished_constructions     | 0.1519                                |
| google_trends              | 0.9313                                |

From the results we see a strong positive relationship between the house price index and the google trend one,a more moderate one with the shipping price index and the fx indicator. While the weaker relationships are found with the energy price index and the unemployment rate. These correlation serve as a point of reference for the linear regression models that we implement in the corresponding section. Also, we found a high positive correlation between the metal price index and the industrial inputs.

```{r,include=FALSE}
# Identify columns that do not contain "log"
columns_to_keep <- !grepl("log", names(data_reduced))

# Subset the dataframe to keep only these columns
df_not_log <- data_reduced[, columns_to_keep]

# Remove y_1
#df_not_log <- subset(df_not_log, select = -y_1)

# Display the filtered dataframe
print(df_not_log)
numeric_columns <- select_if(df_not_log, is.numeric)
```

Revisar si tiene sentido esta matris
```{r,include=FALSE}
cov_matrix <- cov(numeric_columns)
cov_matrix
```


```{r,include=FALSE}
correl_matrix <- cor(numeric_columns)
correl_matrix
# Heatmap of correl
```


```{r,echo=FALSE,fig.width=7, fig.height=7}
{image(1:ncol(correl_matrix), 1:ncol(correl_matrix), correl_matrix,
      main = "Correlation Matrix Heatmap",
      xlab = "Variables", ylab = "Variables", axes = FALSE,
      col = heat.colors(20))
axis(1, at = 1:ncol(correl_matrix), labels = colnames(correl_matrix), las = 2)
axis(2, at = 1:ncol(correl_matrix), labels = colnames(correl_matrix), las = 2)}
```

# Log-transformation

We apply a log transformation with the intention of normalize the data as some variables are index that change over time and others area real values in a wide range. Our idea is to reduce the variability of the data, compressing the scale and helping to linearize those relationships between the variables for the interpretability of the results in the models. We apply log transformation to all variables except those that are already percentages: the unemployment rate and the interest rate.
```{r}
#Log transformation of the variables
variables_to_log <- c("Home_Price_Index", "Industrial_Inputs_Index", "Metals_Price_Index",
                      "Energy_Price_Index", "Shipping_Price_Index", "Forex_Index",
                      "Industrial_Production_Index", "Construction_Licences_Area",
                      "Finished_Constructions", "Google_Trends_Housing")


for (var in variables_to_log) {
  log_var_name <- paste0(var, "_log")
  data_reduced[[log_var_name]] <- log(data_reduced[[var]])
}

# Subset the dataframe
selected_columns <- c(paste0(variables_to_log, "_log"), "Unemployment_Rate", "Interest_Rate")
data_final <- data_reduced[, selected_columns]

```


# Multicolinearity Check

Now present a multicollinearity test in top of the results related with the correlation of the variables in relation with the dependent variable of House Price Index. We first calculate the VIF for each predictor variable. We found that some variables like Shipping, Industrial Production Index, Metals Price Index and Industrial Inputs present a high VIF. 

```{r,echo=FALSE}
# here a value bigger than 10 is supposed to have high multicolinearity 
vif_values <-vif(lm(Home_Price_Index_log ~ ., data = data_final))
vif_values
# some variables like shipping, industrial prod, metals and industrial input are high,
# Metals and industrial inputs are very similar / metals is a part of industrial inputs,
# drop metals
```
At this point we consider to remove the Metals Price Index as they show a strong correlation with Industrial Inputs, they both have a similar behavior and at the end the metal sector is consider on the Industrial Inputs Index. After this reduction we re run the VIF test. EN ESTE PUNTO HABLAR DE POR QUÉ DEJAR ALGUNAS AUNQUE TENGA VIF ALTO. 

```{r,echo=FALSE}
data_final$Metals_Price_Index_log<- NULL

# re run multicoliniarity

vif(lm(Home_Price_Index_log ~ ., data = data_final))

# Industrial production and unemployment could be related but cant delete one at this point.

# consider regularization using ridge or lasso
# or see how it behaves without industrial inputs

# Assume var_to_exclude is the name of the variable you want to exclude
# var_to_exclude <- "industrial_inputs_log"
# 
# # Create a formula that excludes the specific variable
# predictors <- setdiff(names(data_final), c("y_log", var_to_exclude))
# formula <- as.formula(paste("y_log ~", paste(predictors, collapse = " + ")))
# 
# # Fit the model and calculate VIF
# vif_values <- vif(lm(formula, data = data_final))
# print(vif_values)
```
# Train Test Split

```{r}

# # Sample data preparation
# set.seed(123)  # For reproducibility
# trainIndex <- createDataPartition(data_final$Home_Price_Index_log, p = .8, 
#                                   list = FALSE, 
#                                   times = 1)
# data_train <- data_final[ trainIndex,]
# data_test  <- data_final[-trainIndex,]
# 
# # Extract predictors and response
# X_train <- model.matrix(Home_Price_Index_log ~ ., data_train)[, -1]  # Remove the intercept column
# y_train <- data_train$Home_Price_Index_log
# 
# X_test <- model.matrix(Home_Price_Index_log ~ ., data_test)[, -1]  # Remove the intercept column
# y_test <- data_test$Home_Price_Index_log

# Sample data preparation
set.seed(123)  # For reproducibility

# Create a sample index for training data (80% of the data)
trainIndex <- sample(seq_len(nrow(data_final)), size = 0.8 * nrow(data_final))

# Split the data into training and testing sets
data_train <- data_final[trainIndex, ]
data_test <- data_final[-trainIndex, ]

# Extract predictors and response for the training set
X_train <- model.matrix(Home_Price_Index_log ~ ., data = data_train)[, -1] 
y_train <- data_train$Home_Price_Index_log

# Extract predictors and response for the testing set
X_test <- model.matrix(Home_Price_Index_log ~ ., data = data_test)[, -1] 
y_test <- data_test$Home_Price_Index_log
```


# Modelling the data

We start with a simple regression model for the data, remember that this process will be executed with the variables after the log normalization. In this first iteration we consider all the explanatory variables.

```{r}
# OLS 1 Log-Log-----------------
# Perform linear regression
model <- lm(Home_Price_Index_log ~ ., data = data_train)  
summary(model)
```

In this first model we see a high R-squared, this mean that the model is powerful in terms of explain the variance in the House Price Index, but the model is no completely satisfactory as it includes several predictors without statistical significance, and other with different degree of signifance as industrial production, google trends and interest rate. This can be explain because the R-squared statistic will always increase when more variables are added to the model.

```{r,include=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(model, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(model, which = 2, main = "Normal Q-Q")
}

```

After having the model we fit it to evaluate its performance. First we determine the MSE which has a value of 0.001106908 suggesting the the predictions are close to the actual values, this could be observed in the plot of the predicted vs the actual values which presents an approximately 45° degrees line.

```{r, echo=FALSE}
# Predict and evaluate the model on the test set
predictions <- predict(model, newdata = data_test)
actuals <- data_test$Home_Price_Index_log

# Calculate evaluation metrics
mse <- mean((predictions - actuals)^2)
print(mse)
plot(actuals,predictions)
```

Then we plot the residuals Histogram and QQ plot. The histogram shows an approximately normally distribution centered around zero. However, the QQ plot show some deviations outside the reference line.





```{r,include=FALSE}
# # Do Test of normality to residuals
# 
# # For all these tests, the null hypothesis H0  is that the residuals are normally distributed.
# # pvalue > 0.05 -> Fail to reject the null hypothesis, suggesting that the residuals are normally distributed.
# # Perform Shapiro-Wilk test
# shapiro_test <- shapiro.test(residuals)
# print(shapiro_test)
# 
# # Perform Kolmogorov-Smirnov test
# ks_test <- ks.test(residuals, "pnorm", mean = mean(residuals), sd = sd(residuals))
# print(ks_test)
# 
# # Both p-values are > 0.05, so the residuals should be normal
# 
# 
# # Multicolinearity
# 
# vif(model)
# # Shipping and Industrial production could be linearly correlated
# 
# 
# # Model fit
# summary(model)$r.squared
# 
# 
# # Influential plots
# # Cook's Distance Plot
# {plot(cooks.distance(model), main="Cook's Distance")
# abline(h = 4/(nrow(data_final)-length(coef(model))), col="red")}  # Common threshold line
# 
# # Leverage Plot
# {plot(hatvalues(model), main="Leverage Values")
# abline(h = 2*mean(hatvalues(model)), col="red")}  # Common threshold line
# 

```

# Drop variables according to their significance

After the first iteration we consider the process of eliminating variables based on its significance to the model. The first variable that we consider to discard is the Energy Price Index, for this we proceed with an ANOVA F-test for the comparison of regression models. As the obtained value is 0.1328 is greater than 0.05 the Energy Price Index is not a significant predictor in the presence of the other variables. With this we remove it

```{r,echo=FALSE}
# Variable Selection-----------
# F-test for the comparison of nested models
full.mod <- lm(Home_Price_Index_log ~ ., data = data_train)
summary(full.mod)
# Remove energy
red.mod <- update(full.mod, . ~ . -Energy_Price_Index_log)
anova(red.mod, full.mod)
```

We observe again the significance of the variables, and the next no significant variables that we decide to drop were Industrial Inputs Index, Finished Constructions, Construction Licence Area and Unemployment. All of this after eliminating step by step each one following the previous procedure with the ANOVA F-test until all the predictors are significative. Justificar porque dejamos constructions licences aunque no es significativa

```{r,results='hide'}
#Process of Variable dropping according to ther significance

# Energy does not improve model
summary(red.mod)
red.mod2 <- update(red.mod, . ~ . -Industrial_Inputs_Index_log)
anova(red.mod, red.mod2)
# Can safely remove industrial inputs log
summary(red.mod2)

# Remove finished_constructions_log (only price variable so far)
red.mod3 <- update(red.mod2, . ~ . -Finished_Constructions_log)
anova(red.mod2, red.mod3)
# Can safely remove finished constructions (but makes no sense)
summary(red.mod3)
# All predictors are statistically significant

#red.mod4 <- update(red.mod3, . ~ . -Construction_Licences_Area_log)
#anova(red.mod3, red.mod4)
# Can safely remove constructions_licences_area_log
#summary(red.mod4)


# All predictors are statistically significant
```

```{r,echo=FALSE}
summary(red.mod3)
```



```{r,include=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(red.mod3, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(red.mod3, which = 2, main = "Normal Q-Q")
}

```
The final reduced model continue to show a high R-squared, while the QQ-Plot shows more points outside the reference line. At this point even the fact that now all the variables are significative the model appears to has less explanatory power as the histogram of the residuals show a more skewed behavior even though it is centered in zero.

```{r, echo=FALSE}
# Predict and evaluate the model on the test set
predictions <- predict(red.mod3, newdata = data_test)
actuals <- data_test$Home_Price_Index_log

# Calculate evaluation metrics
mse <- mean((predictions - actuals)^2)
print(mse)
plot(actuals,predictions)
```



# Correction of the model with lagged value
```{r}
# Using a a lag version of the Home Price Index

data_final_with_lag<- data_final

data_final_with_lag$Home_Price_Index_log_lag1 <- c(NA, head(data_final$Home_Price_Index_log, -1))

# Remove rows with NA values
data_final_with_lag <- na.omit(data_final_with_lag)

# Fit the linear regression model

```

```{r}
# Sample data preparation
# set.seed(123)  # For reproducibility
# 
# trainIndex_lag <- createDataPartition(data_final_with_lag$Home_Price_Index_log, p = .8, 
#                                   list = FALSE, 
#                                   times = 1)
# data_train_lag <- data_final_with_lag[ trainIndex_lag,]
# data_test_lag  <- data_final_with_lag[-trainIndex_lag,]
# 
# # Extract predictors and response
# X_train_lag <- model.matrix(Home_Price_Index_log ~ ., data_train_lag)[, -1]  # Remove the intercept column
# y_train_lag <- data_train_lag$Home_Price_Index_log
# 
# X_test_lag <- model.matrix(Home_Price_Index_log ~ ., data_test_lag)[, -1]  # Remove the intercept column
# y_test_lag <- data_test_lag$Home_Price_Index_log

# Set the seed for reproducibility
set.seed(123)

# Assuming data_final_with_lag is your dataframe and Home_Price_Index_log is your response variable

# Create a sample index for training data (80% of the data)
trainIndex_lag <- sample(seq_len(nrow(data_final_with_lag)), size = 0.8 * nrow(data_final_with_lag))

# Split the data into training and testing sets
data_train_lag <- data_final_with_lag[trainIndex_lag, ]
data_test_lag <- data_final_with_lag[-trainIndex_lag, ]

# Extract predictors and response for the training set
X_train_lag <- model.matrix(Home_Price_Index_log ~ ., data = data_train_lag)[, -1]  # Remove the intercept column
y_train_lag <- data_train_lag$Home_Price_Index_log

# Extract predictors and response for the testing set
X_test_lag <- model.matrix(Home_Price_Index_log ~ ., data = data_test_lag)[, -1]  # Remove the intercept column
y_test_lag <- data_test_lag$Home_Price_Index_log
```

As we continue tosee some variability in the residuals, we decided to use a lagged version of the dependent variable as an explanatory variable. This can help to account for the lagged effects between the variables.

```{r}

model_lag <- lm(Home_Price_Index_log ~ Home_Price_Index_log_lag1 +Forex_Index_log+Industrial_Production_Index_log+Construction_Licences_Area_log+Google_Trends_Housing_log+Interest_Rate, data = data_train_lag)

# Summarize the model
summary(model_lag)
# red.trial <- update(model_lag, . ~ . -)
# anova(model_lag, red.trial)

#We drop Unemployment,Shipping_Price 
```

After include the lagged effect we consider the reduced model showing a high R-squared. Then we proceed to show the histogram of the residuals and their QQ plot. From the histogram it is possible to see that now the residuals show a more


```{r,include=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(model_lag, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(model_lag, which = 2, main = "Normal Q-Q")
}

```

```{r, echo=FALSE}
# Predict and evaluate the model on the test set
predictions <- predict(model_lag, newdata = data_test_lag)
actuals <- data_test_lag$Home_Price_Index_log

# Calculate evaluation metrics
mse <- mean((predictions - actuals)^2)
print(mse)
plot(actuals,predictions)
```



# Forward AIC Selection

We Implement the forward selection based on the AIC obtained by the scores. The AIC approach allows a model comparison with the intention to find a good balance between the model fit and the complexity. The process of forward selection stars with no predictors and adding one by one. In thes case we have a no significative predictor the Industrial Inputs Index, this happens because with the AIC criterion is possible to obtain non-significant predictors if their inclusion improves the general performance of the model.

```{r,results='hide'}
# Forward AIC Stepwise selection-------------

# Fit the null model (intercept only)
null_model <- lm(Home_Price_Index_log ~ 1, data = data_train)

# Specify the full model with all potential predictors
full_model <- lm(Home_Price_Index_log ~ Industrial_Inputs_Index_log + Shipping_Price_Index_log + Forex_Index_log + Industrial_Production_Index_log + 
                   Construction_Licences_Area_log + Google_Trends_Housing_log + 
                   Unemployment_Rate + Interest_Rate, data = data_train)

# Perform forward selection based on AIC
forward_model <- step(null_model, 
                      scope = list(lower = null_model, upper = full_model), 
                      direction = "forward")
```


```{r,echo=FALSE}
# Display the summary of the selected model
summary(forward_model)
```

```{r,include=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(forward_model, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(forward_model, which = 2, main = "Normal Q-Q")
}

```

```{r, echo=FALSE}
# Predict and evaluate the model on the test set
predictions <- predict(forward_model, newdata = data_test)
actuals <- data_test$Home_Price_Index_log

# Calculate evaluation metrics
mse <- mean((predictions - actuals)^2)
print(mse)
plot(actuals,predictions)
```

# Backward AIC selection

We implement the backward selection based also based in the AIC obtained by the different models. In this case, contrary to forward selection, all the variables left were in the reduced model obtained by leaving only the most significative variables. so for this case the results are the same obtained before.

```{r,results='hide'}
# Backward AIC Stepwise selection-------------

full_model <- lm(Home_Price_Index_log ~ Industrial_Inputs_Index_log + Shipping_Price_Index_log + Forex_Index_log + Industrial_Production_Index_log + 
                   Construction_Licences_Area_log + Google_Trends_Housing_log + 
                   Unemployment_Rate + Interest_Rate, data = data_train)

# Perform backward selection based on AIC
backward_model <- step(full_model, 
                       direction = "backward")
```


```{r,echo=FALSE}
# Display the summary of the selected model
summary(backward_model)

# Backward gives construction licences, which is good

# Overall the models converge to this last model (removing by p-value), stepwise...
# So we can do the post estimation of the red.mod3, because is the same,
# but use the coefficients of the model with robust errors for inference
```
El modelo es igual al forward entonces no le hacemos nada

```{r,include=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(backward_model, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(backward_model, which = 2, main = "Normal Q-Q")
}

```



# Lasso regresion model

```{r}
# Regularization------------
# Done over full model
# Load the packages
library(glmnet)



# Fit LASSO regression model
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1)  # alpha = 1 for LASSO
lasso_best_lambda <- lasso_model$lambda.min
lasso_predictions <- predict(lasso_model, s = lasso_best_lambda, newx = X_test)
actuals <- y_test
```





```{r}
# Coefficients at best lambda
lasso_coefficients <- coef(lasso_model, s = lasso_best_lambda)
print(lasso_coefficients)
```

```{r, echo=FALSE,fig.align='center'}
lasso_residuals <- y_test - lasso_predictions
fitted_values <- predict(lasso_model, s = lasso_best_lambda, newx = X_train)
residuals <- y_train - fitted_values
{
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
# Calculate the residuals for the training set
# Plot Residuals vs Fitted Values
plot(fitted_values, residuals, main = "Residuals vs Fitted Values",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

qqnorm(lasso_residuals)
qqline(lasso_residuals, col = "red")}
```

```{r}
# Calculate the Mean Squared Error (MSE)
mse <- mean((lasso_predictions - y_test)^2)
print(paste("Mean Squared Error:", mse))

# Calculate R-squared (R2)
y_mean <- mean(y_test)
ss_total <- sum((y_test - y_mean)^2)
ss_residual <- sum((y_test - lasso_predictions)^2)
r2 <- 1 - (ss_residual / ss_total)
print(paste("R-squared:", r2))
{
plot(actuals,predictions)
abline(0, 1, col = "red")}
```




# Ridge regression model
```{r}
# Regularization------------
# Done over full model
# Load the packages
library(glmnet)



# Fit RIDGE regression model
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0)  # alpha = 1 for LASSO
ridge_best_lambda <- ridge_model$lambda.min
ridge_predictions <- predict(ridge_model, s = ridge_best_lambda, newx = X_test)
actuals <- y_test
```


```{r}
# Coefficients at best lambda
ridge_coefficients <- coef(ridge_model, s = ridge_best_lambda)
print(ridge_coefficients)
```


```{r, echo=FALSE,fig.align='center'}
ridge_residuals <- y_test - ridge_predictions
fitted_values <- predict(ridge_model, s = ridge_best_lambda, newx = X_train)
residuals <- y_train - fitted_values
{
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
# Calculate the residuals for the training set
# Plot Residuals vs Fitted Values
plot(fitted_values, residuals, main = "Residuals vs Fitted Values",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

qqnorm(ridge_residuals)
qqline(ridge_residuals, col = "red")}
```


```{r}
# Calculate the Mean Squared Error (MSE)
mse <- mean((ridge_predictions - y_test)^2)
print(paste("Mean Squared Error:", mse))

# Calculate R-squared (R2)
y_mean <- mean(y_test)
ss_total <- sum((y_test - y_mean)^2)
ss_residual <- sum((y_test - ridge_predictions)^2)
r2 <- 1 - (ss_residual / ss_total)
print(paste("R-squared:", r2))
{
plot(actuals,predictions)
abline(0, 1, col = "red")}
```


