---
title: "statistical_project"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2024-05-27"
Authors: Daniel Gutierrez, Rafael Hernandez, Camilo Sanabria
---

# Project Introduction

In this project we aim to study the influence of selected macroeconomics indicators on the house price index in Colombia. This indicator measure the evolution through time of the market prices of residential properties. The idea of the analysis is to present and study the possible effects of the macroeconomics indicators on the house prices through the construction of regression models, evaluating their results and interpreting the models in the context of the data. This will help understand the main price drivers and could eventually help policy makers to provide an ecosystem for home affordability.

# Dataset Introduction and Preprocessing

The dataset was constructed gathering information from multiple sites: the Colombian department of statistics (DANE) , the Colombian central bank (Banco de la República), Google trends queries and the Federal Reserve Economic Data (FRED). In general, all the visited sites present the data as a .csv file with tables according to filters related with the time span of interest. With this we construct a consolidated database in .csv format with the following variables:

| Variable                   | Description                                                                                                    |
|------------------------|------------------------------------------------|
| House Price Index       | Measure of the average change in the residential properties price as an index                                  |
| Industrial Inputs Index          | Measure of the average changes in the industry input costs as an index                                         |
| Metals Price Index                     | Measure of the average price change in metal costs excluding gold as an index                                  |
| Energy Price Index                     | Measure of the average price change of energy including Crude oil, Natural Gas, Coal Price and Propane Indices |
| Shipping Price Index                   | Measure the average price change of shipping costs                                                             |
| Forex Index                         | Indicator of the foreign exchange rate of the Colombian Peso (COP) with respect of the USD                     |
| Unemployment Rate              | Indicator of the percentage of unemployment population in Colombia                                             |
| Industrial Production Index           | Measure of the level of production on industrial sectors as an index                                           |
| Interest rate              | Reference interest rate emited by the colombian central bank with respect of the others financial institutions |
| Construction Licences Area | Measure the total of squared meters given for construction licences in Colombia                                |
| Finished Constructions     | Measure of the total constructions that have been finished in Colombia                                         |
| Google Trends Housing              | Google search trends on housing for Colombia                                                                   |

These variables present real values,and were selected as they have an initial coherent relation with the housing sector. Initially  We present the import process of the data
```{r,echo=FALSE}
#`echo = FALSE`: Hides the R code in the final document.
#`results = 'hide'`: Hides the output of the R code.
```

```{r import-data, echo=FALSE,message=FALSE, warning=FALSE}
# Necessary libraries to run all the code chunks and plots
library(readxl)
library(car) 
library(lmtest) 
library(zoo)
#library(caret)
```

# Data Uploading

```{r}
#Read the data using the read_excel function
#Change here for current data path
data <- read_excel("C:/Users/danie/Documents/GitHub/unipd_sl_24/data/data.xlsx",sheet = "dataframe_col")
```

# Data pre-processing and cleaning

After have an initial version of the data we proceed with the pre-processing and cleaning steps. As  firs step we change the original variable names for more appropriate ones. Then we identify the number of NA values in each one of the columns.

```{r}
#change the variables names to have more consistency 
colnames(data) <- c('MY','date','House_Price_Index','Industrial_Inputs_Index','Metals_Price_Index','Energy_Price_Index','Shipping_Price_Index','Forex_Index','Unemployment_Rate','Industrial_Production_Index','Interest_Rate','Construction_Licences_Area','Finished_Constructions','Google_Trends_Housing') 
# Counting NA values in each column
na_counts <- apply(data, 2, function(x) sum(is.na(x)))
# Print the counts of NA values per column
print(na_counts)
# Convert date column to Date type if it's not already
data$date <- as.Date(data$date)
# Copy the dataframe to avoid modifying the original one
data_filled <- data
```
We realize that the variable Finished_Constructions requires an additional pre-processing step because it is recorded by quarters while the others are registered by months. We propose the following reconstruction steps for those intermediate months. The idea is to take the quarterly value and divide it equally among the respective months of that quarter:

```{r}
# Find the indices where the date is at the end of a trimester
trimester_end_indices <- which(format(data$date, "%m") %in% c("03", "06", "09", "12"))
# Loop through each trimester end distribute the value
for (i in trimester_end_indices) {
  if (i - 2 > 0) {
    # Distribute the value to the current month and the previous two months
    value_to_distribute <- data$Finished_Constructions[i] / 3
    data_filled$Finished_Constructions[i] <- value_to_distribute
    data_filled$Finished_Constructions[i - 1] <- value_to_distribute
    data_filled$Finished_Constructions[i - 2] <- value_to_distribute
  }
}
data_filled$Finished_Constructions <- na.locf(data_filled$Finished_Constructions, na.rm = FALSE)
```


# Summary stats and final dataset

We proceed to remove the left rows with NA values eliminating from the dataset the registers before January 2005. Also, we drop the column MY also referred to the date in form of MONTH()YEAR(). Finally, we present an initial summary of the dataset after these transformations. In particular we see that the House Price Index range present values on [87.35,203.50] with an IQR=70.38.

```{r}
# Removing rows with any NA values
clean_data <- na.omit(data_filled)
# Drop MY column
data_reduced <- subset(clean_data, select = -c(MY))
```


```{r,results = 'hide'}
summary_stats <- summary(data_reduced[, !(names(data_reduced) %in% c("date"))])
```


| Statistic    | Min    | 1st Qu. | Median | Mean   | 3rd Qu. | Max    |
|--------------|--------|---------|--------|--------|---------|--------|
| House_Price_Index            | 87.35  | 110.50  | 157.92  | 147.71 | 180.88  | 203.50 |
| Industrial_Inputs_Index     | 78.67  | 116.02  | 136.26  | 137.78 | 157.65  | 221.14 |
| Metals_Price_Index          | 74.89  | 121.21  | 143.74  | 145.88 | 172.53  | 234.53 |
| Energy_Price_Index          | 55.89  | 128.19  | 162.61  | 173.36 | 223.35  | 376.41 |
| Shipping_Price_Index        | -213.3 | 242.2   | 262.1   | 280.7  | 294.1   | 478.3  |
| Forex_Index                 | 0.1712 | 1.1934  | 2.3762  | 2.6762 | 3.2021  | 4.9223 |
| Unemployment_Rate           | 0.07563| 0.09728 | 0.10970 | 0.11240| 0.12099 | 0.21972|
| Industrial_Production_Index | 59.82  | 77.94   | 94.44   | 92.87  | 104.85  | 132.46 |
| Interest_Rate               | 0.01750| 0.03860 | 0.04525 | 0.05551| 0.06994 | 0.13250|
| Construction_Licences_Area  | 233210 | 969559  | 1156424 | 1205587| 1373585 | 2883878|
| Finished_Constructions      | 471310 | 775374  | 848663  | 871488 | 962168  | 1335811|
| Google_Trends_Housing       | 16.00  | 23.50   | 34.75   | 36.04  | 47.62   | 64.00  |

# Exploratory Analysis


# Outliers based on the IQR

As an initial step to better understand the data and its behavior we find the outliers of each one of the variables using the IQR to select those points outside the bounds. As a result we found that the variables that present the most quantity of outliers are the Shipping Price Index, the Interest Rate and the Unemployment Rate. In addition, we present the boxplot of these three variables to obtain a clear summary of their distribution.  

```{r,results = 'hide'}
# Function to detect outliers based on IQR
detect_outliers_single_var <- function(column) {
  column <- na.omit(column) 
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  outliers <- column[column < lower_bound | column > upper_bound]
  return(outliers)
}
# List to store outliers for each variable
outliers_list <- list()
# Loop to determine outliers in each column
for (var in colnames(data_reduced[, !(names(data_reduced) %in% c("date"))])) {
  if (is.numeric(data[[var]])) {
    outliers_list[[var]] <- detect_outliers_single_var(data[[var]])
  }
}
print(outliers_list)
```
| Variable                      | Values                                                                                                     |
|-------------------------------|------------------------------------------------------------------------------------------------------------|
| House_Price_Index              | 0                                                                                                |
| Industrial_Inputs_Index       | 0                                                                                                |
| Metals_Price_Index            | 0                                                                                                |
| Energy_Price_Index            | 376.4121                                                                                                  |
| Shipping_Price_Index          | 374.969, 375.187, 395.268, 446.983, 439.181, 454.407, 468.721, 478.263, 478.216, 471.877, 469.290, 442.591, 456.584, 412.082, 409.599, 443.799, 420.694, 446.214, 462.462  |
| Forex_Index                   | 0                                                                                                |
| Unemployment_Rate             | 0.1725640, 0.2048530, 0.2197200, 0.2035910, 0.2091470, 0.1744270, 0.1629280, 0.1756237                    |
| Industrial_Production_Index   | 0                                                                                                |
| Interest_Rate                 | 0.1141935, 0.1204839, 0.1275000, 0.1275806, 0.1300000, 0.1324194, 0.1325000, 0.1325000                    |
| Construction_Licences_Area    | 2493085, 1988396, 2838778, 2337743, 2331210, 2859035, 2372414, 2052607                                    |
| Finished_Constructions        | 4007432, 3762696, 1413931                                                                                 |
| Google_Trends_Housing         | 0        

```{r,echo = FALSE,fig.width=3, fig.height=5,fig.align='center'}
#Create box plot
{
par(mfrow = c(3, 1), mar = c(5, 4, 2, 2) + 0.1, oma = c(0, 0, 0, 0))
boxplot(data_reduced$Shipping_Price_Index, horizontal=TRUE  ,frame=F,main = "Shipping Price Index")
boxplot(data_reduced$Unemployment_Rate, horizontal=TRUE  ,frame=F,main = "Unemployment Rate")
boxplot(data_reduced$Interest_Rate, horizontal=TRUE  ,frame=F,main = "Interest Rate")
}
```
The outlier data for the shipping index, unemployment, and interest rates all relate to the COVID-19 shock. The shipping index crisis in mid-2022 resulted from the COVID-19 lockdowns (2020-2021), economic reactivation, and delays in Chinese and Western ports. The unemployment outliers are due to the initial COVID lockdown in Colombia and the ensuing economic shock. The high interest rates stem from the Colombian Central Bank's efforts to control inflation starting in 2021.

# Ploting the House Price Index

Continuing with the exploratory analysis we plot the variable of House Price Index to see its behavior through time. It is possible to see a clear increase tendency starting from the year 2005 until the year 2021, then a little decay was recorded.

```{r,echo = FALSE,fig.width=6, fig.height=4,fig.align='center'}
par(mar = c(5, 4, 4, 2) + 0.1)
par(mfrow = c(1, 1))
# dependent variable
data_reduced$category <- as.factor(clean_data$date)
# Set up the plot without drawing the points
{plot(1:length(data_reduced$category), clean_data$House_Price_Index, type = "n", xaxt = "n",xlab="date", ylab = "House Price Index", main = "Evolution of House Prices over time",cex.axis=0.8)
# Add the categorical labels to the x-axis
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)

# Add the lines and points to the plot
lines(1:length(data_reduced$category), data_reduced$House_Price_Index, type = "o", col = "blue", pch = 16)}
```

# Ploting percentual variation of the House Price Index

Further useful information can be found in the box plots and histograms for the percentual variations. We take the 12-period percentual variation of the variables with the intention of identify patterns and tendencies in a clear and interpretable way, considering that the majority of the variables are in index form, and the study of their changes is fundamental for a complete understanding of their behavior through time. We start with the House Price Index variable.

In these plots we see that the central value of the percentual variations is positive, indicating and overall increase in the House Price Index through time. Also the negatives outliers were expected because those negatives variations due to the decrease after the 2021. In the histogram it is possible to see a right skewed behavior, give by precisely those increments until the 2021.

```{r}
# Select subset of variables to difference
variables_to_calculate <- c("House_Price_Index", "Industrial_Inputs_Index", "Metals_Price_Index",
                            "Energy_Price_Index", "Shipping_Price_Index", "Forex_Index",
                            "Industrial_Production_Index", "Construction_Licences_Area",
                            "Finished_Constructions", "Google_Trends_Housing")

# Create a function to calculate the 12-month percentual variation
percentual_variation_12_months <- function(x) {
  return((x / lag(x, n = 12) - 1))
}
lag <- function(x, n) {
  c(rep(NA, n), head(x, -n))
}
data_percentual_variation <- data
# Apply the function to the select variables
for (var in variables_to_calculate) {
  new_var_name <- paste0("pct_var_", var)
  data_percentual_variation[[new_var_name]] <- percentual_variation_12_months(data[[var]])
}

```

```{r,echo = FALSE,fig.width=9, fig.height=4,fig.align='center'}
# Histograms/Boxplots of the percentual variations
# Subset only percentual changes
columns_to_select <- c("pct_var_House_Price_Index", "pct_var_Industrial_Inputs_Index", "pct_var_Metals_Price_Index", 
                       "pct_var_Energy_Price_Index", "pct_var_Shipping_Price_Index", "pct_var_Forex_Index", 
                       "pct_var_Industrial_Production_Index", "pct_var_Construction_Licences_Area", 
                       "pct_var_Finished_Constructions", "pct_var_Google_Trends_Housing", 
                       "Unemployment_Rate", "Interest_Rate")

data_percentual_variation <- data_percentual_variation[, columns_to_select]

data_percentual_variation$time_index<- seq(1,nrow(data_percentual_variation))


{

par(mar = c(5, 4, 4, 2) + 0.1)
par(mfrow = c(1, 2))

# Boxplot with color scale for time index
boxplot(data_percentual_variation$pct_var_House_Price_Index, horizontal = TRUE, frame = FALSE, 
        main = "Boxplot of Percentual Variation House Price Index", 
        ylab = "Percentual Variation", xlab = "", col = "lightgray", border = "black")

# Generate color scale based on time index
colors <- colorRampPalette(c("green", "red"))(length(data_percentual_variation$time_index))
colored_points <- colors[rank(data_percentual_variation$time_index)]

# Overlay points on the boxplot
points(data_percentual_variation$pct_var_House_Price_Index, rep(1, nrow(data_percentual_variation)), 
       col = colored_points, pch = 19)

# Add a legend for the color scale
legend("topright", legend = c("Oldest", "Newest"), 
       fill = c(colors[1], colors[length(colors)]), title = "Time Index")

# Histogram of Percentual Variation House Price Index
hist(data_percentual_variation$pct_var_House_Price_Index, breaks = 40, 
     main = "Histogram of Percentual Variation House Price Index", 
     xlab = "Percentual Variation")}
```

We did a similar procedure for the rest of the variables with respect to their percentual variation. For an easier interpretation we plot the histograms for the percentual variation of each variable. This plot shows that the index variables with a high variability in percentual variation are the shipping index price and the finished constructions. Also, the index variable that show less variability in the percentual variation is the one related with the industrial production.
```{r}
{par(mfrow = c(3,4 ), mar = c(2, 2, 2, 2) + 0.1)
data_percentual_variation <- data_percentual_variation[, columns_to_select]
for (i in 1:ncol(data_percentual_variation)) {
  hist(data_percentual_variation[[i]], 
       main = substr(names(data_percentual_variation)[i], 1, 25), # Shorten title if necessary
       xlab = names(data_percentual_variation)[i], 
       col = "blue", 
       border = "black")
}
}
```

```{r,echo = FALSE,fig.width=6, fig.height=6}
# This is to plot the boxplot, we dont use it becouse it requires the shape library
#par(mfrow = c(1, 1))
# 
#pairs(data_percentual_variation)
# 
# #Boxplot for all at once
# #install.packages("reshape2")
# library(reshape2)
# # Melt the dataframe to long format
# # Melt the dataframe to long format
# data_long <- melt(data_percentual_variation)
# 
# # Adjust graphical parameters
# 
# 
# # Create the boxplot without x-axis labels
# {par(mar = c(5, 5, 4, 2) + 0.1)  
#   boxplot(value ~ variable, data = data_long,
#         main = "Boxplot of Multiple Variables",
#         xlab = "", ylab = "Value",
#         col = "lightblue", border = "darkblue", xaxt = 'n')
# 
# # Add custom x-axis labels at a 45-degree angle
# labels <- levels(data_long$variable)
# text(x = 1:length(labels), y = par("usr")[3] - 0.5, labels = labels, srt = 45, adj = 1, xpd = TRUE, cex = 0.8)
# }
# 
# # Reset graphical parameters to default
# #par(mar = c(5, 4, 4, 2) + 0.1)
# # This one goes in the report, not so much variability in prices even a lot in X vars
# 

```

We plot these four variables to see its evolution through time. It is possible to see that the industrial production Index and the Shipping price Index have an increase tendency, with some peaks as for example the negative peak related with the COVID pandemic for the Industrial Production Index on the 2020. In another hand, the Finished Constructions have a less defined behavior over time.

```{r,echo = FALSE,fig.width=3, fig.height=6,fig.align='center'}
# reset margins

# dependent variable
data_reduced$category <- as.factor(clean_data$date)
# Set up the plot without drawing the points
{par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(3, 1))
  plot(1:length(data_reduced$category), clean_data$Shipping_Price_Index, type = "n", xaxt = "n",xlab="date", ylab = "Shipping Price Index", main = "Evolution of Shipping Prices Index over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Shipping_Price_Index, type = "o", col = "blue", pch = 16)

plot(1:length(data_reduced$category), clean_data$Finished_Constructions, type = "n", xaxt = "n",xlab="date", ylab = "Energy Price Index", main = "Evolution of Finished constructions over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Finished_Constructions, type = "o", col = "blue", pch = 16)

plot(1:length(data_reduced$category), clean_data$Industrial_Production_Index, type = "n", xaxt = "n",xlab="date", ylab = "Industry Production Index", main = "Evolution of Industrial Production Index over time")
axis(1, at = 1:length(data_reduced$category), labels = data_reduced$category)
lines(1:length(data_reduced$category), data_reduced$Industrial_Production_Index, type = "o", col = "blue", pch = 16)
}
```

# Log-transformation

We apply a log transformation with the intention of scaling the data as some variables are index that change over time and others area real values with wider range. Our goal is to reduce the variability of the data, compressing the scale and helping to linearize those relationships between the variables for the interpretability of the results in the subsequent models. We apply log transformation to all variables except those that are already percentages: the Unemployment Rate and the Interest Rate.

```{r}
#Log transformation of the variables
variables_to_log <- c("House_Price_Index", "Industrial_Inputs_Index", "Metals_Price_Index",
                      "Energy_Price_Index", "Shipping_Price_Index", "Forex_Index",
                      "Industrial_Production_Index", "Construction_Licences_Area",
                      "Finished_Constructions", "Google_Trends_Housing")


for (var in variables_to_log) {
  log_var_name <- paste0(var, "_log")
  data_reduced[[log_var_name]] <- log(data_reduced[[var]])
}

# Subset the dataframe
selected_columns <- c(paste0(variables_to_log, "_log"), "Unemployment_Rate", "Interest_Rate")
data_final <- data_reduced[, selected_columns]

```

# Variable Correlation

Now we present the  correlation between the variables of the dataset in order to understand the strength of their relation. We present the correlation matrix obtained. From the results we see a strong positive relationship between the House Price Index and the Google Trend and Industrial Production one,a more moderate correlation with the shipping Price and the Forex Index. While the weaker relationships are found with the Energy Price index and the Unemployment Rate.For the others variables, we found a high positive correlation between the metal price index and the industrial inputs.

```{r,include=FALSE}
cov_matrix <- cov(data_final)
cov_matrix
```


```{r,include=FALSE}
correl_matrix <- cor(data_final)
correl_matrix
print(correl_matrix)
```

```{r}
# Plot the correlation matrix 

{par(mar = c(7, 7, 7, 2) + 0.1)
cor_matrix<-correl_matrix
image(1:ncol(cor_matrix), 1:nrow(cor_matrix), t(cor_matrix)[, ncol(cor_matrix):1], 
      axes = FALSE, xlab = "", ylab = "", col = colorRampPalette(c("red", "white", "blue"))(20))

# Add axis labels 
axis(3, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 2, cex.axis = 0.7)
axis(2, at = 1:nrow(cor_matrix), labels = rev(rownames(cor_matrix)), las = 2, cex.axis = 0.7)

#correlation coefficients
for (i in 1:ncol(cor_matrix)) {
  for (j in 1:nrow(cor_matrix)) {
    text(i, nrow(cor_matrix) - j + 1, round(cor_matrix[i, j], 2), cex = 0.7)
  }
}}
```

# Multicolinearity Check

Now present a multicollinearity test in top of the results related with the correlation of the variables in relation with the dependent variable of House Price Index. We first calculate the VIF for each predictor variable. We found that some variables like Shipping, Industrial Production Index, Metals Price Index and Industrial Inputs present a high VIF. 

```{r,echo=FALSE}
# here a value bigger than 10 is supposed to have high multicolinearity 
vif_values <-vif(lm(House_Price_Index_log ~ ., data = data_final))
vif_values
# some variables like shipping, industrial prod, metals and industrial input are high,
# Metals and industrial inputs are very similar / metals is a part of industrial inputs,
# drop metals
```
At this point we consider to remove the Metals Price Index as they show a strong correlation with Industrial Inputs, they both have a similar behavior and at the end the metal sector is considered on the Industrial Inputs Index. After this reduction we re run the VIF test. There are some variables that show high VIF, but due to the economic literature we decide to keep going with them. Later we will perform variable selection through different approaches and eventually the multicolinearity here present will be corrected.

```{r,echo=FALSE}
data_final$Metals_Price_Index_log<- NULL

# re run multicoliniarity

vif(lm(House_Price_Index_log ~ ., data = data_final))

```
# Train Test Split
Before proceed with the linear models we apply a train test split in order to have a test set to compare the performance in a fair way between the data. For this we implement the following steps, considering that our train set will be 80% of the data: 
```{r}
set.seed(123)  
# sample index for training data
trainIndex <- sample(seq_len(nrow(data_final)), size = 0.8 * nrow(data_final))
# Split the data into training and testing sets
data_train <- data_final[trainIndex, ]
data_test <- data_final[-trainIndex, ]

X_train <- model.matrix(House_Price_Index_log ~ ., data = data_train)[, -1] 
y_train <- data_train$House_Price_Index_log
X_test <- model.matrix(House_Price_Index_log ~ ., data = data_test)[, -1] 
y_test <- data_test$House_Price_Index_log
```


# Modelling the data

#Initial full linear model 
We start with a simple regression model for the data, remember that this process will be executed with the variables after the log normalization. In this first iteration we consider all the explanatory variables.


```{r}
# OLS 1 Log-Log-----------------
# Perform linear regression
model <- lm(House_Price_Index_log ~ ., data = data_train)  
summary(model)
```

In this first model we see a high R-squared, this mean that the model is powerful in terms of explain the variability in the House Price Index, but the model is no completely satisfactory as it includes several predictors without statistical significance, and other with different degree of significance as Industrial Production Index, Google Trends Housing and Interest Rate. This can be explain because the R-squared statistic will always increase when more variables are added to the model. We proceed to plot the the results for the fitted results and the QQ-plot of the residuals:

```{r,echo=FALSE,fig.width=7, fig.height=4}
{par(mfrow = c(1, 2))
# Plot residuals vs fitted values
plot(model, which = 1, main = "Residuals vs Fitted")
# Plot QQ plot
plot(model, which = 2, main = "Normal Q-Q")
}

```

After having the model we fit it to evaluate its performance. First we determine the MSE which has a value of 0.00269 suggesting the the predictions are close to the actual values, this could be observed in the plot of the predicted vs the actual values which presents an approximately 45° degrees line.

```{r, echo=FALSE,fig.width=4, fig.height=3}
# Predict and evaluate the model on the test set
predictions <- predict(model, newdata = data_test)
actuals <- data_test$House_Price_Index_log

# Calculate evaluation metric mse
mse_full <- mean((predictions - actuals)^2)
print(mse_full)
{plot(predictions, actuals, main = "Prediction vs Actual Values full model",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0,1, col = "red")}
```


# Model dropping variables according to their significance

After the first iteration we consider the process of eliminating variables based on its significance to the model. The first variable that we consider to discard is the Energy Price Index, for this we proceed with an ANOVA F-test for the comparison of regression models. As the obtained value is 0.1328 is greater than 0.05 the Energy Price Index is not a significant predictor in the presence of the other variables. With this we remove it

```{r,echo=FALSE}
# Variable Selection-----------
# F-test for the comparison of nested models
full.mod <- lm(House_Price_Index_log ~ ., data = data_train)
summary(full.mod)
# Remove energy
red.mod <- update(full.mod, . ~ . -Energy_Price_Index_log)
anova(red.mod, full.mod)
```

We observe again the significance of the variables, and the next no significant variables that we decide to drop were Industrial Inputs Index, Finished Constructions. All of this after eliminating step by step each one following the previous procedure with the ANOVA F-test until all the predictors show statistical evidence of having coefficient different from 0.

```{r,results='hide'}
#Process of Variable dropping according to ther significance

# Energy does not improve model
summary(red.mod)
red.mod2 <- update(red.mod, . ~ . -Industrial_Inputs_Index_log)
anova(red.mod, red.mod2)
# Can safely remove industrial inputs log
summary(red.mod2)

# Remove finished_constructions_log 
red.mod3 <- update(red.mod2, . ~ . -Finished_Constructions_log)
anova(red.mod2, red.mod3)
# Can safely remove finished constructions (but makes no sense)
summary(red.mod3)
```

```{r,echo=FALSE}
summary(red.mod3)
```



```{r,echo=FALSE,fig.width=7, fig.height=4}

{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(red.mod3, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(red.mod3, which = 2, main = "Normal Q-Q")
}

```
The final reduced model continue to show a high R-squared, while the QQ-Plot shows similar points outside the reference line. At this point even the fact that now all the variables are statistically significant, the model appears to has more or less the same explanatory power due to the residuals behavior.

```{r, echo=FALSE,fig.width=4, fig.height=3}
# Predict and evaluate the model on the test set
predictions <- predict(red.mod3, newdata = data_test)
actuals <- data_test$House_Price_Index_log

# Calculate evaluation metric mse
mse_red <- mean((predictions - actuals)^2)
print(mse_red)
{plot(predictions, actuals, main = "Prediction vs Actual Values significance",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0,1, col = "red")}
```


# Model with lagged variable

As a following step we implement a procedure to consider a lag version of the House Price Index, this due to the possible influence of lagged affects, in terms of the intuition that the actual house price index will not be determined by the current explanatory variables, but by the explanatory variables of the previous period. Research conducted by Iacoviello, Matteo (2004) suggests that the house prices can be modeled with a VAR approach, which considers lags of both the dependent variable and the covariates. We include so far only the dependent variable due to the modelling approach in this written (Linear Regression Analysis). Unemployment Rate was discarded because its coefficient showed no evidence of being different from zero, and via the ANOVA analysis it could be concluded that it was safe to discard it.
```{r}
# Using a a lag version of the House Price Index

data_final_with_lag<- data_final

data_final_with_lag$House_Price_Index_log_lag1 <- c(NA, head(data_final$House_Price_Index_log, -1))

data_final_with_lag <- na.omit(data_final_with_lag)


```

```{r}

# Set the seed for reproducibility
set.seed(123)

trainIndex_lag <- sample(seq_len(nrow(data_final_with_lag)), size = 0.8 * nrow(data_final_with_lag))

# Split the data into training and testing sets
data_train_lag <- data_final_with_lag[trainIndex_lag, ]
data_test_lag <- data_final_with_lag[-trainIndex_lag, ]


X_train_lag <- model.matrix(House_Price_Index_log ~ ., data = data_train_lag)[, -1] 
y_train_lag <- data_train_lag$House_Price_Index_log


X_test_lag <- model.matrix(House_Price_Index_log ~ ., data = data_test_lag)[, -1]  
y_test_lag <- data_test_lag$House_Price_Index_log
```

```{r}

model_lag <- lm(House_Price_Index_log ~ House_Price_Index_log_lag1 +Forex_Index_log+Industrial_Production_Index_log+Construction_Licences_Area_log+Google_Trends_Housing_log+Interest_Rate, data = data_train_lag)

# Summarize the model
summary(model_lag)

```

After include the lagged effect we consider the reduced model showing a high R-squared. Then we proceed to show the histogram of the residuals and their QQ plot. From the QQ-plot it is possible to see that now the residuals show a even strong normal behavior


```{r,echo=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(model_lag, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(model_lag, which = 2, main = "Normal Q-Q")
}

```

Now, it is possible to see that the fit of the predictions and actual values is more correct, in terms that it follows almost a a perfect straight line.  

```{r, echo=FALSE,,fig.width=4, fig.height=3}
predictions <- predict(model_lag, newdata = data_test_lag)
actuals <- data_test_lag$House_Price_Index_log

# Calculate evaluation metric mse
mse_lag <- mean((predictions - actuals)^2)
print(mse_lag)
{plot(predictions, actuals, main = "Prediction vs Actual Values lagged model",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0,1, col = "red")}
```



# Forward AIC Selection model

We Implement the forward selection based on the AIC for a family of models. The AIC approach allows a model comparison with the aim to find a good balance between model fit and complexity. The process of forward selection starts with no predictors (just with the bias) and adds one by one each of the covariates from the pre-cured dataframe. We could see variables with no statistical evidence of significance using this approach. This happens because via the AIC forward selection method is possible to obtain  non-significant coefficients for a given set of variables but their inclusion minimizes the AIC.

```{r,results='hide'}
# Forward AIC Stepwise selection-------------

# Fit the null model (intercept only)
null_model <- lm(House_Price_Index_log ~ 1, data = data_train)

# Specify the full model with all potential predictors
full_model <- lm(House_Price_Index_log ~ Industrial_Inputs_Index_log + Shipping_Price_Index_log + Forex_Index_log + Industrial_Production_Index_log + 
                   Construction_Licences_Area_log + Google_Trends_Housing_log + 
                   Unemployment_Rate + Interest_Rate, data = data_train)

# Perform forward selection based on AIC
forward_model <- step(null_model, 
                      scope = list(lower = null_model, upper = full_model), 
                      direction = "forward")
```


```{r,echo=FALSE}
# Display the summary of the selected model
summary(forward_model)
```

```{r,echo=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(forward_model, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(forward_model, which = 2, main = "Normal Q-Q")
}

```

```{r, echo=FALSE,,fig.width=4, fig.height=3}
# Predict and evaluate the model on the test set
predictions <- predict(forward_model, newdata = data_test)
actuals <- data_test$House_Price_Index_log

# Calculate evaluation metric mse
mse_forward <- mean((predictions - actuals)^2)
print(mse_forward)
{plot(predictions, actuals, main = "Prediction vs Actual Values Forward Seelction",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0,1, col = "red")}
```

# Backward AIC selection model

We also implemented backward selection based on the AIC obtained from various models. We start from the full model then variables are removed based on the AIC. As a result, the outcomes matched those previously obtained in the forward selection approach.

```{r,results='hide'}
# Backward AIC Stepwise selection-------------

full_model <- lm(House_Price_Index_log ~ Industrial_Inputs_Index_log + Shipping_Price_Index_log + Forex_Index_log + Industrial_Production_Index_log + 
                   Construction_Licences_Area_log + Google_Trends_Housing_log + 
                   Unemployment_Rate + Interest_Rate, data = data_train)

# Perform backward selection based on AIC
backward_model <- step(full_model, 
                       direction = "backward")
```


```{r,echo=FALSE}
# Display the summary of the selected model
summary(backward_model)
```


```{r,echo=FALSE,fig.width=7, fig.height=4}
# Post estimation Model 1----------------
# See standard R plots
{par(mfrow = c(1, 2))

# Plot residuals vs fitted values
plot(backward_model, which = 1, main = "Residuals vs Fitted")

# Plot QQ plot
plot(backward_model, which = 2, main = "Normal Q-Q")
}

```



# Lasso regresion model

As an alternative approach, we decided to implement the Lasso regression model. Our goal is to obtain a model that shrinks the coefficients of the variables, thereby reducing overfitting. Another aim is to evaluate whether some coefficients of the explanatory variables are set to zero. To achieve this, we use cross-validation to train the model.  

```{r}

library(glmnet)

# Fit LASSO regression model
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1)  # alpha = 1 for LASSO
lasso_best_lambda <- lasso_model$lambda.min
lasso_predictions <- predict(lasso_model, s = lasso_best_lambda, newx = X_test)
actuals <- y_test
```
We obtained the Lasso coefficients and observed that none were shrunk to zero. However, some coefficients were significantly reduced, indicating a lower influence on the House Price Index. For example, the coefficients for Finished Constructions and the Energy Price Index were notably shrunk by this regularization technique.
```{r}
# Coefficients at best lambda
lasso_coefficients <- coef(lasso_model, s = lasso_best_lambda)
print(lasso_coefficients)
```

```{r, echo=FALSE,fig.align='center',fig.width=7, fig.height=4}
lasso_residuals <- y_test - lasso_predictions
fitted_values <- predict(lasso_model, s = lasso_best_lambda, newx = X_train)
residuals <- y_train - fitted_values
{
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
# Calculate the residuals for the training set
# Plot Residuals vs Fitted Values
plot(fitted_values, residuals, main = "Residuals vs Fitted Values",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

qqnorm(lasso_residuals)
qqline(lasso_residuals, col = "red")}
```

```{r,fig.width=4, fig.height=3}
# Calculate the Mean Squared Error (MSE)
mse_lasso <- mean((lasso_predictions - y_test)^2)
print(paste("Mean Squared Error:", mse_lasso))

# Calculate R-squared (R2)
y_mean <- mean(y_test)
ss_total <- sum((y_test - y_mean)^2)
ss_residual <- sum((y_test - lasso_predictions)^2)
r2 <- 1 - (ss_residual / ss_total)
print(paste("R-squared:", r2))
{
plot(actuals,predictions,main = "Prediction vs Actual Values lasso",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0, 1, col = "red")}
```
# Ridge regression model

We applied a similar procedure for Ridge regression to assess whether it produces a different model compared to Lasso, given its distinct regularization term. The results are presented above:


```{r}
# Regularization------------
# Done over full model

# Fit RIDGE regression model
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0)  # alpha = 0 for Ridge
ridge_best_lambda <- ridge_model$lambda.min
ridge_predictions <- predict(ridge_model, s = ridge_best_lambda, newx = X_test)
actuals <- y_test
```


```{r}
# Coefficients at best lambda
ridge_coefficients <- coef(ridge_model, s = ridge_best_lambda)
print(ridge_coefficients)
```
In general we get a different results for the coefficients with Ridge. It is possible to see that some of them change drastically as for example the one for the Unemployment Rate.In the Lasso approach Unemployment Rate has a positive high value, while with Ridge we obtained a negative small one. According to economic theory, the sign of Unemployment should be negative, because as Unemployment goes up, there should be less available income for the families, reducing the demand for new homes, thus having a negative impact on the final price of new homes. So the Ridge coefficient makes more sense.
```{r, echo=FALSE,fig.align='center',fig.width=7, fig.height=4}
ridge_residuals <- y_test - ridge_predictions
fitted_values <- predict(ridge_model, s = ridge_best_lambda, newx = X_train)
residuals <- y_train - fitted_values
{
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
# Calculate the residuals for the training set
# Plot Residuals vs Fitted Values
plot(fitted_values, residuals, main = "Residuals vs Fitted Values",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

qqnorm(ridge_residuals)
qqline(ridge_residuals, col = "red")}
```


```{r,,fig.width=4, fig.height=3}
# Calculate the Mean Squared Error (MSE)
mse_ridge <- mean((ridge_predictions - y_test)^2)
print(paste("Mean Squared Error:", mse_ridge))

# Calculate R-squared (R2)
y_mean <- mean(y_test)
ss_total <- sum((y_test - y_mean)^2)
ss_residual <- sum((y_test - ridge_predictions)^2)
r2 <- 1 - (ss_residual / ss_total)
print(paste("R-squared:", r2))
{
plot(actuals,predictions,main = "Prediction vs Actual Values Ridge",
     xlab = "Fitted Values", ylab = "Predictions")
abline(0, 1, col = "red")}
```
#MSE Model comparison


| Variable  | Value |
|-----------|-------|
| mse_full | 1.867e-3    |
| mse_red | 1.830e-3    |
| mse_lag | 0.0231e-3    |
| mse_forward | 1.830e-3    |
| mse_backward | 1.830e-3    |
| mse_lasso | 1.832e-3    |
| mse_ridge | 2.096e-3    |


It is evident that the best model in terms of MSE is the one using the lagged dependent variable as a predictor. Compared to the others, most models provide similar values for the coefficients of the explanatory variables. In general, the reduced model obtained with only the statistically significant variables and the one implemented using AIC-based selection perform similarly. The Lasso and Ridge regularization techniques yield similar results, with Ridge having a slightly higher MSE but more reasonable signs for its coefficients.


# Statistical Inference - Best Model


Based on the model with the lowest MSE, we now proceed to interpret the coefficients of the covariates. All variables have a p-value indicating statistical significance at least at the 5% level. The log-transformation of the variables that were not percentages is handy as it allows us to interpret the log-log effects as the impact of a 1% increase in variable X on the House Price Index by B%.

In this model, the variable with the highest coefficient is the lagged House Price Index. This aligns with the economic theory of price stickiness and the inflationary spiral. A 1% increase in the House Price Index results in a 0.96% increase for the next month, indicating high inflationary persistence. It is important to note that durable goods tend to have stickier prices. The second coefficient we observe is for Forex, showing that a 1% increase in the foreign exchange rate reduces the House Price Index by 0.009%. This coefficient is almost negligible, which makes sense since the instantaneous effect on house prices is minimal.

Next, we observe the Industrial Production Index, where a 1% increase results in a 0.03% increase in the House Price Index. This makes sense as a more dynamic economy typically leads to higher demand for housing, thus increasing prices. For the Construction Licenses Area, a 1% increase results in a 0.003% decrease in the House Price Index. This aligns with economic theory, where an increase in available housing area should lead to an increase in supply, thereby reducing prices.

The Google Trends Index shows an effect of 0.008%, indicating that the intention to buy a house, or at least the related searches, positively impacts the dependent variable. This is notable because this indicator can be obtained in real time, unlike the others and can help predict the direction of housing prices in advance. Lastly, for the Interest Rate, a 1% increase leads to a 0.07% decrease in prices. This is because higher interest rates make consumers less likely to take on credit, thereby reducing the demand for new houses and consequently lowering prices.

For further studies, different lags for the explanatory variables could be added to determine which have the most impact on the House Price Index. While this is beyond the scope of this report, it could be useful for government policy-making based on statistical evidence. This way, housing prices can be managed more effectively, contributing to the long-term wealth of the population.

